---
title: "s04_add_phenotypes"
output: html_document
editor_options: 
  chunk_output_type: console
---

started: AL10Sep2018  
last updated: AL30Sep2019  

# Summary  

- Initially, phenotype data were provided by Xiaolin in e-mail to Marc of 29Aug2018  
  (used here in file phenotypes_dec2018.txt)  
- Then, updated data were provided to me by Marc in e-mail of 07Jan2019  
  (file phenotypes_jan2019.txt, contains data about familial history, X-ray treatment etc)  
- The script reads both sets and compares them before use  
- The script syncs phenotype cases IDs with the samples names in sequencing 
- Then script renames and syncronises data (removes cases failed sequencing from phenotype data)   

Input and Output data: 3,758 vars x 535 samples (265UBC + 270CBC)  

# start_section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
#options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F,
        warnPartialMatchArgs = T, 
        warnPartialMatchAttr = T, 
        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_wo_danish_2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s09_add_phenotypes")
#dir.create(data_folder)
scripts_folder <- file.path(base_folder,"scripts","s09_add_phenotypes")
setwd(scripts_folder)

# Libraries
library(dplyr)

```

# Load data

## Phenotype data with Danish  

```{r}

load(file.path(base_folder,"data","s05_phenotypes","misc","add_ampliseq_danish_samples.RData"))

# Update folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_wo_danish_2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s09_add_phenotypes")
scripts_folder <- file.path(base_folder,"scripts","s09_add_phenotypes")

# Clean-up
rm(working_folder)

```

## Phenotype data w/o Danish  

```{r}

phenotypes_2018_file <- file.path(data_folder, "phenotypes_txt","phenotypes_dec2018.txt")
phenotypes_2018.df <- read.table(phenotypes_2018_file, header=T, sep="\t", quote="")

phenotypes_2019_file <- file.path(data_folder, "phenotypes_txt", "phenotypes_jan2019.txt")
phenotypes_2019.df <- read.table(phenotypes_2019_file, header=T, sep="\t", quote="")

# Clean-up
rm(phenotypes_2018_file, phenotypes_2019_file)

```

# Sequencing data

```{r}

load(file.path(base_folder,"data","s08_update_variant_annotations","s03_update_variant_annotations.RData"))

# Update folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_wo_danish_2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s09_add_phenotypes")
scripts_folder <- file.path(base_folder,"scripts","s09_add_phenotypes")

```

# Compare phenotypes provided in 2018 and 2019

All fields from 2018 are included in 2019  
Additional fields in 2019: family_history, setno, xrtbreast  

```{r}

dim(ampliseq_samples.df)
dim(phenotypes_2018.df)
dim(phenotypes_2019.df)

colnames(ampliseq_samples.df)
colnames(phenotypes_2018.df)
colnames(phenotypes_2019.df)

all(phenotypes_2018.df == ampliseq_samples.df[,1:18])
all(phenotypes_2018.df == phenotypes_2019.df[,1:18])

identical_fields <- colnames(phenotypes_2018.df)
sum(phenotypes_2018.df[,identical_fields] != phenotypes_2019.df[,identical_fields])

sum(is.na(phenotypes_2018.df))
sum(is.na(phenotypes_2019.df))


rm(identical_fields)

```

# Rename matrices

```{r update_variants}

variants.df <- vars.df
genotypes.mx <- gt.mx
phenotypes.df <- phenotypes_2019.df

sum(is.na(phenotypes.df))
sum(phenotypes.df=="")

dim(phenotypes.df)
dim(variants.df)
dim(genotypes.mx)

rm(gt.mx, vars.df, phenotypes_2018.df, phenotypes_2019.df)

```

# Add wecare_phenotypes IDs  

The initial phenotype table includes samples numbers (e.g. 40),  
while the sequencing data include alphanumerical IDs (e.g. 40_S189_L007)  

This chunk adds alphanumerical IDs to the phenotype table as rownames  

```{r phenotypes_rownames}

# Get list of long IDs for all 541 samples
vcf_file="/Users/alexey/Documents/wecare/ampliseq/v04_ampliseq_nfe/s04_annotated_vcf/ampliseq_nfe.vcf"
vcfr <- read.vcfR(vcf_file)
gt_739.mx <- extract.gt(vcfr)

dim(gt_739.mx)
gt_739.mx[1:5,1:5]
gt_739.mx[1:5,541:542]
gt_739.mx[1:5,735:739]
wecare_samples.df <- data.frame(long_ids=colnames(gt_739.mx)[1:541])
nfe_samples.df <- data.frame(long_ids=colnames(gt_739.mx)[542:739])

# Split the long IDs for wecare samples
wecare_samples.df <- separate(wecare_samples.df, "long_ids", c("num", "illumina_id", "illumina_lane"), 
                              sep="_", remove=F) 
wecare_samples.df$num <- as.integer(wecare_samples.df$num)
dim(wecare_samples.df)
str(wecare_samples.df)
wecare_samples.df[1:5,]

# Sort wecare_samples.df by sample num (to sync with phenotypes.df)
wecare_samples.df <- wecare_samples.df[order(wecare_samples.df$num),]
wecare_samples.df[1:5,]
phenotypes.df[1:5,1:5]

# Bind add long IDs to phenotypes table
phenotypes.df <- cbind(wecare_samples.df, phenotypes.df)

# Check that numbers are identical
sum(phenotypes.df$num == phenotypes.df$Sample_num)
sum(phenotypes.df$num != phenotypes.df$Sample_num)

# Remove one of the numbers
phenotypes.df <- phenotypes.df[,-2]
dim(phenotypes.df)
colnames(phenotypes.df)
str(phenotypes.df)

# Set rownames
rownames(phenotypes.df) <- phenotypes.df$long_ids

# Check results
genotypes.mx[1:5,1:5]
variants.df[1:5,1:5]
phenotypes.df[1:5,1:5]

dim(genotypes.mx)
dim(variants.df)
dim(phenotypes.df)

# Clean-up
rm(vcf_file, vcfr, gt_739.mx, wecare_samples.df, nfe_samples.df)

```

# Sync phenotypes and genotypes

Remove samples failed sequencing (6 wecare and 1 nfe)  

```{r sync_pheno_geno}

dim(genotypes.mx)
dim(phenotypes.df)

phenotypes.df <- phenotypes.df[colnames(genotypes.mx),]

dim(phenotypes.df)

```

# check data

```{r end_check}

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

sum(colnames(genotypes.mx) != rownames(phenotypes.df))
sum(rownames(genotypes.mx) != rownames(variants.df))

table(phenotypes.df$cc)

```

# save_results

```{r save_results}

save.image(paste(base_folder, "s01_add_phenotypes.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
