---
title: "Compare phenotypes of 2018 and 2019"
author: "AL"
date: "03 Mar 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
editor_options:
  chunk_output_type: console
---

# Summary

- Initially, Ampliseq phenotype data were provided by Xiaolin in e-mail to Marc of **29Aug2018**  
  file name: **Phenotype data for exome seq_shipping manifest_sent_10 05 17.xlsx**  
- Then, the updated data were provided to me by Marc in e-mail of **07Jan2019**  
  file name: **2Phenotype data for exome seq_shipping manifest_sent_10 05 17.xlsx**  

This script reads both sets and compares them before use.  
The comparison shows that data of 2019 is identical to data of 2018, except for addition of 3 new columns:  
- family_history  
- setno  
- xrtbreast  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F)

#options(warnPartialMatchArgs = T, 
#        warnPartialMatchAttr = T, 
#        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_wo_danish_2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s09_add_phenotypes")
#dir.create(data_folder)
scripts_folder <- file.path(base_folder,"scripts","s09_add_phenotypes")
setwd(scripts_folder)

# Libraries
library(XLConnect)

```

# Read source data

## 2018

```{r}

data_file_2018 <- file.path(base_folder,"data","s05_phenotypes","Phenotype data for exome seq_shipping manifest_sent_10 05 17.xlsx")

wb_2018 <- loadWorkbook(data_file_2018)

phenotypes_2018.df <- readWorksheet(wb_2018, sheet = "Phenotype data")
dim(phenotypes_2018.df)
colnames(phenotypes_2018.df)

rm(data_file_2018, wb_2018)

```

## 2019

```{r}

data_file_2019 <- file.path(base_folder,"data","s05_phenotypes","2Phenotype data for exome seq_shipping manifest_sent_10 05 17.xlsx")

wb_2019 <- loadWorkbook(data_file_2019)

phenotypes_2019.df <- readWorksheet(wb_2019, sheet = "Phenotype data")
dim(phenotypes_2019.df)
colnames(phenotypes_2019.df)

rm(data_file_2019, wb_2019)

```

# Compare data

```{r}

all(phenotypes_2018.df == phenotypes_2019.df[,1:18])

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
