---
title: "Add flag for Danish ampliseq samples"
author: "AL"
date: "03 Mar 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: 
      collapsed: false
      smooth_scroll: false
editor_options:
  chunk_output_type: console
---

# Summary

- The updated phenotype data were provided to me by Marc in e-mail of **07Jan2019**  
    + file name: **2Phenotype data for exome seq_shipping manifest_sent_10 05 17.xlsx**  
- The information about Danish samples was sent by Xiaolin to Marc on **11Sep2019**  
    + file name: **Danish_samples_in_WES.xlsx**

In e-mail of **27Apr2020** Marc clarified that clinically actionable genes must not be analysed in Danish cohort.  
So, these samples should be excluded from the current analysis because it aggregates variants in the known BC genes.  

This script flags Danish samples in ampliseq phenotype data.  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F)

#options(warnPartialMatchArgs = T, 
#        warnPartialMatchAttr = T, 
#        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_wo_danish_2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s09_add_phenotypes")
#dir.create(data_folder)
scripts_folder <- file.path(base_folder,"scripts","s09_add_phenotypes")
setwd(scripts_folder)

# Libraries
library(XLConnect)
library(dplyr)

```

# Read source data

## 2019  

```{r}

# Read
data_file_2019 <- file.path(base_folder,"data","s05_phenotypes","2Phenotype data for exome seq_shipping manifest_sent_10 05 17.xlsx")
wb_2019 <- loadWorkbook(data_file_2019)
phenotypes_2019.df <- readWorksheet(wb_2019, sheet = "Phenotype data")

# Explore
dim(phenotypes_2019.df)
colnames(phenotypes_2019.df)
phenotypes_2019.df[1:5,1:5]

# Clean-up
rm(data_file_2019, wb_2019)

```

```{r}

# Read
data_file <- file.path(base_folder,"data","s05_phenotypes","Danish_samples_in_WES.xlsx")
wb <- loadWorkbook(data_file)
danish_samples.df <- readWorksheet(wb,sheet = "DE_WES_ROUND2")

# Explore
dim(danish_samples.df)
colnames(danish_samples.df)
danish_samples.df[c(1,38,74),]

# Clean-up
rm(data_file,wb)

```

# Add danish information

```{r}

danish <- phenotypes_2019.df$Sample_num %in% danish_samples.df$Sample_number
sum(danish)

phenotypes_2019_with_danish.df <- data.frame(phenotypes_2019.df,danish)
str(phenotypes_2019_with_danish.df)
sum(phenotypes_2019_with_danish.df$danish)

rm(danish_samples.df, phenotypes_2019.df, danish)

```

# Save result

```{r}

ls()
save.image(file.path(data_folder,"s02_flag_danish_samples.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
