---
title: "Filter variants ampliseq"
author: "Alexey Larionov"
date: "27 Feb 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary 

- keep only FILTER == PASS  
- Explore QUAL and DP  
- Additionally: remove ad.mx  

Input data: 12,276 variants x 541 cases  
Output data: 7,762 vars x 541 cases  

# start_section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r start_section}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F,
        warnPartialMatchArgs = T, 
        warnPartialMatchAttr = T, 
        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_wo_danish_2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s07_qc_filters")
dir.create(data_folder)
scripts_folder <- file.path(base_folder,"scripts","s07_qc_filters")
setwd(scripts_folder)

```

# read_data  

```{r load_data}

load(file.path(base_folder,"data","s06_read_data","s02_remove_nfe.RData"))

rm(ad.mx)

# Update folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_wo_danish_2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s07_qc_filters")
scripts_folder <- file.path(base_folder,"scripts","s07_qc_filters")

```

# check_data

```{r start_check}

# List objects
ls()

# Check sizes
dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)
dim(ref.mx)
dim(alt.mx)

dim(gq.mx)
dim(dp.mx)

dim(fixed.df)
dim(meta.df)

# Check consistence of rownames and colnames

sum(rownames(gt_add.mx) != rownames(gt_num.mx))
sum(rownames(gt_add.mx) != rownames(gt_chr.mx))
sum(rownames(gt_add.mx) != rownames(ref.mx))
sum(rownames(gt_add.mx) != rownames(alt.mx))
sum(rownames(gt_add.mx) != rownames(gq.mx))
sum(rownames(gt_add.mx) != rownames(dp.mx))

sum(colnames(gt_add.mx) != colnames(gt_num.mx))
sum(colnames(gt_add.mx) != colnames(gt_chr.mx))
sum(colnames(gt_add.mx) != colnames(ref.mx))
sum(colnames(gt_add.mx) != colnames(alt.mx))
sum(colnames(gt_add.mx) != colnames(gq.mx))
sum(colnames(gt_add.mx) != colnames(dp.mx))

sum(rownames(gt_add.mx) != rownames(fixed.df))

```

# Explore FILTER

```{r explore_filter}

sort(summary(as.factor(fixed.df$FILTER)), decreasing = T)

```

# Explore QUAL before applying FILTER

```{r qual_before}

hist(fixed.df$QUAL, lab=T, ylim=c(0,15000))
hist(fixed.df$QUAL[fixed.df$QUAL < 100], lab=T, 
     xlim=c(0,100), ylim=c(0,1000))
quantile(fixed.df$QUAL)

```

# Variant DP before applying FILTER

```{r variant_DP_before}

sum(is.na(fixed.df$DP)) # Multiallelic variants

hist(fixed.df$DP, lab=T, ylim=c(0,15000))
hist(fixed.df$DP[fixed.df$DP < 1e5], lab=T, 
     xlim=c(0,1e5), ylim=c(0,2500))
quantile(fixed.df$DP, na.rm=T)

```

# Apply FILTER

```{r apply_filter}

sum(is.na(fixed.df$FILTER))
passed_filter <- fixed.df$FILTER=="PASS"

fixed.df <- fixed.df[passed_filter,]
dim(fixed.df)

gt_add.mx <- gt_add.mx[passed_filter,]
gt_num.mx <- gt_num.mx[passed_filter,]
gt_chr.mx <- gt_chr.mx[passed_filter,]
ref.mx <- ref.mx[passed_filter,]
alt.mx <- alt.mx[passed_filter,]
dp.mx <- dp.mx[passed_filter,]
gq.mx <- gq.mx[passed_filter,]

rm(passed_filter)

```

# QUAL after applying FILTER

```{r qual_after}

hist(fixed.df$QUAL, lab=T, ylim=c(0,10000))
hist(fixed.df$QUAL[fixed.df$QUAL < 100], lab=T, 
     xlim=c(0,100), ylim=c(0,600))
quantile(fixed.df$QUAL)

```

# Variant DP after applying FILTER

```{r variant_DP_after}

sum(is.na(fixed.df$DP)) # Multiallelic variants

hist(fixed.df$DP, lab=T, ylim=c(0,10000))
hist(fixed.df$DP[fixed.df$DP < 1e5], lab=T, 
     xlim=c(0,1e5), ylim=c(0,1500))
quantile(fixed.df$DP, na.rm=T)

```

# check_data

```{r end_check}

# List objects
ls()

# Check sizes
dim(gt_add.mx)
dim(gt_num.mx)
dim(gt_chr.mx)
dim(ref.mx)
dim(alt.mx)

dim(gq.mx)
dim(dp.mx)

dim(fixed.df)
dim(meta.df)

# Check consistence of rownames and colnames

sum(rownames(gt_add.mx) != rownames(gt_num.mx))
sum(rownames(gt_add.mx) != rownames(gt_chr.mx))
sum(rownames(gt_add.mx) != rownames(ref.mx))
sum(rownames(gt_add.mx) != rownames(alt.mx))
sum(rownames(gt_add.mx) != rownames(gq.mx))
sum(rownames(gt_add.mx) != rownames(dp.mx))

sum(colnames(gt_add.mx) != colnames(gt_num.mx))
sum(colnames(gt_add.mx) != colnames(gt_chr.mx))
sum(colnames(gt_add.mx) != colnames(ref.mx))
sum(colnames(gt_add.mx) != colnames(alt.mx))
sum(colnames(gt_add.mx) != colnames(gq.mx))
sum(colnames(gt_add.mx) != colnames(dp.mx))

sum(rownames(gt_add.mx) != rownames(fixed.df))

```

# save_data

```{r save_data}

save.image(file.path(data_folder, "r01_filter_variants.RData"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()
gc()

```
