---
title: "Ampliseq: exclude BRCA1,2 and PALB2 carriers"
author: "Alexey Larionov"
date: "04 Mar 2021"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

# Summary  

Several versions of analysis were previously performed on Ampliseq data, including v3 and v4 of analysis.  These versions resulted to slightly different variants in VCF files because different settings were used (e.g. variant filtering, annotations etc).  All potentially pathogenic QC-passed variants in BRCA1, BRCA2 or PALB2 were manually reviewed in IGV.  

Six BRCA1, BRCA2 or PALB2 carriers suggested by VCF v3 or v4 and confirmed in IGV are excluded by this script:  

- **BRCA1** Var000015470/Var000011226 **48_S416_L008** - detected by both v.3 and v4  
- **BRCA1** Var000011231 **137_S323_L008** - detected by v.4 only  
- **BRCA2** Var000008259 **140_S394_L008** - detected by v.4 only  
- **PALB2** Var000013416 **311_S137_L007** - detected by v.3 only  
- **PALB2** Var000013417 **317_S174_L007** - detected by v.3 only  
- **PALB2** Var000013335 **377_S279_L008** - detected by v.3 only  


Details of BRCA1, BRCA2 or PALB2 carriers from VCF **v.3** (including BAMs & IGV screenshots):  
- /Users/alexey/Documents/wecare/ampliseq/**v03**_ampliseq_only_new/s11_check_BRCA1_BRCA2_PALB2  

Details of BRCA1, BRCA2 or PALB2 carriers from VCF **v.4** (including BAMs & IGV screenshots):  
- /Users/alexey/Documents/wecare/ampliseq/**v04**_ampliseq_nfe/s11_remove_BRCA_PALB_carriers  

46 non-polymorphic sites excluded after the samples filtering  

Input data: 3,721 vars x 521 samples (260UBC + 261CBC)  
Output data: 3,675 vars x 515 samples (258UBC + 257CBC)  

# Start section

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

```{r echo=F}
options(width = 999)
```

```{r}

# Time stamp
Sys.time()

# Clenan-up
rm(list=ls())
graphics.off()

# Memory
gc()

# Options
options(stringsAsFactors = F,
        warnPartialMatchArgs = T, 
        warnPartialMatchAttr = T, 
        warnPartialMatchDollar = T)

# Files and folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_wo_danish_2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s10_filter_samples")
scripts_folder <- file.path(base_folder,"scripts","s10_filter_samples")
setwd(scripts_folder)

```

# Read data  

```{r}

# Load data
load(file.path(data_folder,"s02_remove_duplicates.RData"))

# Update folders
base_folder="/Users/alexey/Documents/wecare/final_analysis_2021/reanalysis_wo_danish_2021/ampliseq/s01_ampliseq_wecare_only"
data_folder <- file.path(base_folder,"data","s10_filter_samples")
scripts_folder <- file.path(base_folder,"scripts","s10_filter_samples")

```

# Check data

```{r}

ls()

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

sum(colnames(genotypes.mx) != rownames(phenotypes.df))
sum(rownames(genotypes.mx) != rownames(variants.df))

```

# List of samples to exclude  

```{r carriers_list}

BRCA_PALB_carriers <- c("48_S416_L008", "137_S323_L008", "140_S394_L008",
                        "311_S137_L007", "317_S174_L007", "377_S279_L008") 


phenotypes.df[BRCA_PALB_carriers,c("cc","danish")]

```

# Exclude the carriers

```{r}

selected_cases <- ! colnames(genotypes.mx) %in% BRCA_PALB_carriers
genotypes.mx <- genotypes.mx[,selected_cases]
phenotypes.df <- phenotypes.df[selected_cases,]

rm(selected_cases, BRCA_PALB_carriers)

```

# Remove non-polymorphic sites

```{r}

# Function to detect uniform numeric vector
uniformity_check.udf <- function(x){
  if (all(is.na(x))){"All_NA"}
  else if (min(x,na.rm=T)==max(x,na.rm=T)){"Uniform"}
  else {"Non-Uniform"}}

# Make the filter
uniformity_check <- apply(genotypes.mx,1,uniformity_check.udf)
summary(as.factor(uniformity_check))
non_uniform_sites <- uniformity_check == "Non-Uniform"

# Remove variants with uniform genotypes accross all samples
genotypes.mx <- genotypes.mx[non_uniform_sites,]
variants.df <- variants.df[non_uniform_sites,]

# Check result
dim(genotypes.mx)
dim(variants.df)

table(phenotypes.df$cc)
table(phenotypes.df$danish)
table(phenotypes.df[,c("cc","danish")])

# Clean-up
rm(uniformity_check.udf, uniformity_check, non_uniform_sites)

```

# Check data

```{r}

ls()

dim(genotypes.mx)
dim(phenotypes.df)
dim(variants.df)

sum(colnames(genotypes.mx) != rownames(phenotypes.df))
sum(rownames(genotypes.mx) != rownames(variants.df))

```

# Save results

```{r}

save.image(file.path(data_folder, "s03_exclude_BRCA1_BCRA2_PALB2_carriers.RData"))

```

# Final section

```{r}

ls()
sessionInfo()
Sys.time()
gc()

```
